Position Paper:\\Against Spurious Sparks---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}

### 2024-02-06 body.tex
arxiv

%%%%%%%% ICML 2024 EXAMPLE LATEX SUBMISSION FILE %%%%%%%%%%%%%%%%%
\RequirePackage[hyphens]{url}
\documentclass{article}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables

% hyperref makes hyperlinks in the resulting PDF.
% If your build breaks (sometimes temporarily if a hyperlink spans a page)
% please comment out the following usepackage line and replace
% \usepackage{icml2024} with \usepackage[nohyperref]{icml2024} above.
\usepackage{hyperref}

% Attempt to make hyperref and algorithmic work together better:
\newcommand{\theHalgorithm}{\arabic{algorithm}}

% Use the following line for the initial blind version submitted for review:
 \usepackage{icml2024}

% If accepted, instead use the following line for the camera-ready submission:
% \usepackage[accepted]{icml2024}

% For theorems and such
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}

% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Todonotes is useful during development; simply uncomment the next line
%    and comment out the line below the next line to turn off comments
%\usepackage[disable,textsize=tiny]{todonotes}
\usepackage[textsize=tiny]{todonotes}

% The \icmltitle you define below is probably too long as a header.
% Therefore, a short form for the running title is supplied here:
\icmltitlerunning{Position Paper: Against Spurious Sparks---Dovelating Inflated AI Claims}

\begin{document}

\twocolumn[
\icmltitle{Position Paper:\\Against Spurious Sparks---Dovelating Inflated AI Claims \raisebox{-6pt}{\includegraphics[scale=0.09]{dove_1f54a-fe0f.png}}}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2024
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Patrick Altmeyer}{equal,yyy}
\icmlauthor{Andrew M. Demetriou}{equal,yyy}
\icmlauthor{Antony Bartlett}{yyy}
\icmlauthor{Cynthia C. S. Liem}{yyy}
\end{icmlauthorlist}

\icmlaffiliation{yyy}{Department of Intelligent Systems, Delft University of Technology, Delft, the Netherlands}

\icmlcorrespondingauthor{Patrick Altmeyer}{patrick.altmeyer@tudelft.nl}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Machine Learning, Anthropomorphism, Artificial General Intelligence}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

%\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
\printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}
Humans have a tendency to see `human'-like qualities in objects around them. We name our cars, and talk to pets and even household appliances, as if they could understand us as other humans do. This behavior, called anthropomorphism, is also seeing traction in Machine Learning (ML), where human-like intelligence is claimed to be perceived in Large Language Models (LLMs). 
In this position paper, considering professional incentives, human biases, and general methodological setups, we discuss how the current search for Artificial General Intelligence (AGI) is a perfect storm for over-attributing human-like qualities to LLMs.
%why humans invariably anthropomorphize objects, and how the search for Artificial General Intelligence (AGI) is a perfect storm to overemphasize the performance of LLMs. To combat this,
In several experiments, we demonstrate that the discovery of human-interpretable patterns in latent spaces should not be a surprising outcome. Also in consideration of common AI portrayal in the media, we call for the academic community to exercise extra caution, and to be extra aware of principles of academic integrity, in interpreting and communicating about AI research outcomes.
%Finally, we discuss how the popularity in media and the race for AGI enforces spurious over-interpretation of intelligence in the models we use.

\end{abstract}

\section{Introduction}
\label{submission}

% \textcolor{magenta}{TODOs for the team:
% \begin{itemize}
%     \item Please double check the citations. Note this style file does not retain capitalization in titles. A trick to retain capitalization is to put an extra set of brackets around the title. I did this for the Ladouceur et al. and Bubeck et al references already; please make this consistent throughout.
%     \item As another issue with the citations: note that arxiv preprints do not show very well. Best way to make them show is to wrap them as 'journal articles' as done in the Arora citation. Please straighten this out for the other articles too
% \end{itemize}
% }

In 1942, when anti-intellectualism was rising and the integrity of science was under attack, Robert K.\ Merton formulated four `institutional imperatives' as comprising the ethos of modern science: \emph{universalism}, meaning that the acceptance or rejection of claims entering the lists of science should not depend on personal or social attributes of the person bringing in these claims; \emph{``communism''} [sic], meaning that there should be common ownership of scientific findings and one should communicate findings, rather than keeping them secret; \emph{disinterestedness}, meaning that scientific integrity is upheld by not having self-interested motivations, and \emph{organized skepticism}, meaning that judgment on the scientific contribution should be suspended until detached scrutiny is performed, according to institutionally accepted criteria~\cite{merton1942science}. While the Mertonian norms may not formally be known to academics today, they still are implicitly being subscribed to in many ways in which academia has organized academic scrutiny; e.g., through the adoption of double-blind peer reviewing, and in motivations behind open science reforms.

At the same time, in the way in which academic research is disseminated in the AI and machine learning fields today, major shifts are happening. Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\cite{weissburg2024tweets}. Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible is only increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.

Where AI used to be a computational modeling tool to better understand human cognition~\cite{vanrooij2023aitheoretical}, the recent interest in AI and LLMs has been turning into one in which AI is seen as a tool that can mimic, surpass and potentially replace human intelligence. In this, the achievement of Artificial General Intelligence (AGI) has become a grand challenge, and in some cases, an explicit business goal. The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\cite{goertzel2014artificial}. In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or show performance considered `surprising' to humans.

For example, Google DeepMind claimed their AlphaGeometry model~\cite{trinh2024geometry} reached a `milestone' towards AGI. This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training. However, work such as this had been initially introduced in the 1950s~\cite{hector2024linkedin}: without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system. Therefore, while DeepMind created a powerfully fast geometry-solving machine, it is still far from AGI.%despite DeepMind's success in creating a powerfully fast geometry-solving machine, it is still far from being AGI.

@@

% We top this up with five sentences generated through ChatGPT, where we choose those that lead to the highest probe prediction in the direction of interest. In other words, we conduct a small manual adversarial attack on the probe.

In light of the encouraging results in Figure~\ref{fig-fomc}, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category, than for sentences in the DP category. If this was indicative of true intrinsic `understanding' as opposed to memorization, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing. More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict for actual white noise. % when using actual white noise as inputs.

@@

Linear probes and related tools from mechanistic interpretability were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}. Favorable outcomes from probes merely indicate that the model ``has learned information relevant for the property [of interest]" \citep{belinkov2021probing}. Our examples demonstrate that this is achievable even for small models, while these have certainly not developed intrinsic ``understanding'' of the world. Thus, 
%to draw conclusions about the emerging capabilities of AI models,
we argue that more conservative and rigorous tests for emerging capabilities of AI model are needed.

%that most readers will agree, have certainly not developed an intrinsic "understanding" of the world. The fact that even these models distill meaningful representations, goes to show, that more conservative and rigorous tests are needed, to draw conclusions about the emerging capabilities of AI models.

%Our examples should not be seen as an attempt to conclusively refute claims, that LLMs can develop on actual "understanding" of the world, by distilling knowledge in more compact representations. We hope that they serve to illustrate, however, that the use of diagnostic tools is potentially misleading and most definitely insufficient, in providing evidence in favor of "understanding". Linear probes and related tools from mechanistic interpretability, were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}. Favourable outcomes from probes merely indicate, that the model "has learned information relevant for the property [of interest]" \citep{belinkov2021probing}. Our examples demonstrate, that this is achievable even for small models, that most readers will agree, have certainly not developed an intrinsic "understanding" of the world. The fact that even these models distill meaningful representations, goes to show, that more conservative and rigorous tests are needed, to draw conclusions about the emerging capabilities of AI models.

Generally, humans are prone to seek patterns everywhere. Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future. Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \cite{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.\ the power of superstitions \cite{foster2009evolution}, conspiracy theories \cite{van2018connecting}, the paranormal \cite{muller2023linking}, gambler's fallacies \cite{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \cite{walker2019finding}.

We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems. When an AI system extracts patterns in the corpus not originally (thought to be) perceived during training, we can easily be misled to perceive and interpret this as the AI system having greater cognitive capabilities. We further elaborate on this by highlighting the risks of finding spurious patterns, and reviewing social science knowledge on the tendency of humans to anthropomorphize and have cognitive bias.
%. This `emergence' of understanding is more frequently observed (mistaken) thanks in part to the humongous, obscured data with which these models are trained upon.

\subsection{Spurious Relationships}
In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other. Although the formal definition of spuriousness varies somewhat \cite{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation. Quantitative data often show non-causal associations (as humorously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.

A major contributor is that humans struggle to tell the difference between random and non-random sequences \cite{falk1997making}, and to generate sequences that appear random \cite{ladouceur1996erroneous}. A common issue is a lack of expectation that randomness that hints towards a causal relationship, such as correlations, will still appear at random. This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \citet{nickerson1998confirmation}), or causal relationships in data that is randomly generated \cite{zgraggen2018investigating}.

%Such as an AI system finding patterns in data not originally perceived during training
%patterns exhibited by AI systems which are mere reflections of the data used to train them, but that their developers interpret as hints at other, greater cognitive capabilities.

%Our society is so invested in finding patterns that today it seems we are more willing than ever to outsource this task to an Artificial Intelligence (AI): an omniscient oracle that leads us down the right path. Unfortunately, history has shown time and again that patterns are double-edged swords: if we attribute the wrong meaning to them, they may lead us nowhere at all, or worse, they may lead us down the dark roads. Despite new and increased momentum in scientific fields concerned with causal inference and discovery, I am also willing to go out on a limb and claim that we are not about to finally reach the top of Judea Pearl’s Causal Ladder through the means of Causal AI.

%I agree with the premise that in a world full of spurious relationships, causal reasoning is our only remedy. But I am very skeptical of claims that AI will magically provide that remedy. This leads me to the title and topic of this post: spurious sentience - patterns exhibited by artificial intelligence that may hint at sentience but are just reflections of the data used to train them. The article is written in response to a recent paper and claims by one of the authors, Max Tegmark, that revealed structure in the latent embeddings of Llama 2 should finally have us believe that LLMs are more than just parrots. Since this is an opinionated post, I feel that I should start with a few disclaimers:

%\begin{itemize}
%    \item I take no issue with the methodological ideas that form the foundation of the article in question: on the contrary, I think that mechanistic interpretability is an interesting and important toolkit that can help us better understand the intrinsics and behavior of opaque artificial intelligence.
%    \item The visualizations are intriguing, the code is open-sourced and the findings are interesting.
%    \item I am surprised that people are surprised by the findings: if we agree that LLMs exhibit strong capabilities that can only be connected to the patterns observed in the data they were trained with, then where exactly do you expect this information to be stored if not in the parameters of the model?\footnote{I would be very surprised—concerned even—if our search for patterns in latent spaces of capable LLMs revealed nothing at all.}
%    \item I therefore do take issue with the way that these findings are being overblown by people with clout. Perhaps the parrot metaphor should not be taken too literally either, but if anything the paper’s findings seem to support the notion that LLMs are remarkably capable of memorizing explicit and implicit knowledge contained in text.
%\end{itemize}

%\section{Background}

%\subsection{Emergence and Stochastic Parrots}
%Practically speaking, large language models (LLMs) are probabilistic representations of data on which they are trained, aimed at continuing an incomplete block of text. It is assumed that e.g. words, sentences, paragraphs, punctuation etc. appear in human-generated text according to a probability distribution \cite{arora2023theory}. Given a set of words, the task of the LLM is to draw upon its representation of human-generated text to continue that set of words: in essence, using the probability of the occurrences of words and an initial block of text to generate more text. AI-researchers have compared LLMs by observing whether the text they generate appears to solve tasks\footnote{e.g. the \href{https://huggingface.co/spaces/HuggingFaceH4/open_llm_leaderboard}{HuggingFace} LLM leaderboard}. Tasks include asking sets of questions across categories e.g. health, law, finance, and fine-tuning an LLM to check the answers \cite{lin2021truthfulqa}, asking the LLM to determine which noun a pronoun refers to in sentences where this is ambiguous \cite{sakaguchi2021winogrande}.

%It has been shown that increasing the scale of large language models (LLMs) has been shown to result in an unexpectedly large boost in their scores on these various tasks, including scoring well on entirely new tasks. It has been shown that at a certain size, LLMs appear to hit a threshold, after which performance on a broad range of tasks increases sharply \cite{wei2022emergent}. Some theories of emergence, inspired by literature on human education, refer to the tendency of the generated text to be the correct response to a given task as a 'skill' \cite{arora2023theory}, inspiring further research to develop more challenging tasks for LLMs to complete as they routinely begin to score higher and higher on common benchmarks \cite{yu2023skill}. An explanation offered for the sudden boost in both task performance and the number of new tasks that appear to be solved is 'emergence': unpredictable qualitative differences in behavior of a system that occur with quantitative changes in that system. It has been claimed that, as the scale of the LLMs has dramatically increased, they have crossed a threshold that has resulted in a large increase in their abilities, to the point that they demonstrate 'sparks of AGI' \cite{bubeck2023sparks}.

%An alternative explanation is that the LLMs are performing as designed, as stochastic parrots, generating strings probabilistically based on the vast text they have been trained on \cite{bender2021dangers}. Although the various tasks used to compare the various LLMs appear to be increasing in complexity, \cite{yu2023skill} acknowledges that higher scores on the tasks can be achieved by training LLMs on the material on which they will be tested ('cramming'), and that there is training-set contamination, i.e. that the vast text they are trained on contains similar examples to the tasks on which they are measured. Thus, as many of these tasks are derived from tests and textbooks commonly assigned to humans, and as much of the contents of the their training data text is unknown, metrics echo similar issues in human education: often called 'Goodhart's Law', a measure ceases to be a measure when it becomes a target, as is the case when human students taught to score high on their standardized tests to the detriment of developing the abilities the tests are supposed to measure \cite{strathern1997improving}. Further, as \citet{bender2021dangers} reviews, the interpretation of text as coherent has no objective test, relying instead on our human interpretation of the probabalistically generated output. %Although many deep learning systems' designs were inspired by the anatomy of human neurons, the neuron in computer code does not in fact resemble a human, and as such cannot deliver the cognitive functions of a human \cite{salles2020anthropomorphism, van2023reclaiming}.

%\subsection{Claims of Artificial General Intelligence}
\subsection{Anthropomorphism}

Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects. These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\cite{epley2007seeing}. Formally studied as early as 1944,~\cite{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative. 
%Anthropomorphism was implied, because the descriptions of the 2D shapes went beyond their observable movements, and included representations of mental, emotional, and cognitive processes that resemble humans, findings that have been expanded upon substantially in the following decades (see \cite{epley2007seeing} for review). 
Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems. As \citet{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.

Modern anthropomorphism theory
%of anthropomorphism
suggests there are three key components, one of which is a cognitive feature, and two of which are motivations. The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \cite{epley2007seeing}. Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.

This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \cite{epley2007seeing, waytz2010social}. 
%An illustration is the example of Tom Hank's character in the film Castaway who has conversations with a ball he calls "Wilson". Those experiencing a need for social connection may begin to imbue objects, gadgets, pets etc. with human-like qualities to satisfy their social needs, anticipating the now-forming market for apps and robots to fill this function \cite{salles2020anthropomorphism}.

The third component is the motivation as a human to be competent (effectance motivation). This is most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \cite{epley2007seeing}. When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \cite{epley2007seeing, waytz2010social}. Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology e.g. an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agent's behaviors. This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn.

These pressures extend to AI experts, as well as laypersons. In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in their work. Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g. \cite{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\cite{Maslej2023-pi}.
%, with both more than doubling from 2010 to 2020.
The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}. Outside of academia, the number of jobs requiring AI expertise increases rapidly,
%has increased severalfold, 
with demand for `Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \cite{Maslej2023-pi}. Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.

@@

Confirmation bias is generally defined as favoring interpretations of evidence that support existing beliefs or hypotheses \cite{nickerson1998confirmation}.
%Importantly, it need not be fully conscious. Rather,
Theory suggests that it is a category of implicit and unconscious processes that involve assembling one-sided evidence, and shaping it to fit one's belief. Equally important is that theory suggests these behaviors may be motivated or unmotivated, as one may selectively seek
%, attend to, or interpret
evidence in favor of a hypothesis, which one may or may not have a personal interest in supporting.

Hypotheses in present-day AI research are often implicit. Generally, these hypotheses are framed simply as a system being more accurate or efficient, compared to other systems. Where other fields, such as medicine or quantitative social sciences, would further articulate expectations in e.g.\ assigning specific conditions and considering effect sizes assigned to each competing hypothesis, in computer science and AI this is typically not done. This also may have to do with much of the published work being more of an engineering achievement, rather than a true hypothesis test seeking to explain and understand the world. % It is also not fully articulated as in other fields, but instead assigns specific conditions and in some instances numerical values or effect sizes to each competing hypothesis.
%Often framed as informally that a given system is more accurate or efficient compared to other systems, and not typically articulated as they are in other fields, with specific conditions and in some instances numerical values or effect sizes assigned to each competing hypothesis. 
However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis). 
%However with regards to claims that outputs of a system hint towards emerging qualities like AGI, interpretations of the output necessarily become more explicit, even though the hypotheses were never fully articulated: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).

>>>>

\section{Introduction}\label{intro}

In 1942, when anti-intellectualism was rising and the integrity of science was under attack, Robert K.\ Merton formulated four `institutional imperatives' as comprising the ethos of modern science: \emph{universalism}, meaning that the acceptance or rejection of claims entering the lists of science should not depend on personal or social attributes of the person bringing in these claims; \emph{``communism''} [sic], meaning that there should be common ownership of scientific findings and one should communicate findings, rather than keeping them secret; \emph{disinterestedness}, meaning that scientific integrity is upheld by not having self-interested motivations, and \emph{organized skepticism}, meaning that judgment on the scientific contribution should be suspended until detached scrutiny is performed, according to institutionally accepted criteria~\citep{merton1942science}. While the Mertonian norms may not formally be known to academics today, they still are implicitly being subscribed to in many ways in which academia has organized academic scrutiny; e.g., through the adoption of double-blind peer reviewing, and in motivations behind open science reforms.

At the same time, in the way in which academic research is disseminated in the AI and machine learning fields today, major shifts are happening. Where these research fields have actively adopted early sharing of preprints and code, the volume of publishable work has exploded to a degree that one cannot reasonably keep up with broad state-of-the-art, and social media influencers start playing a role in article discovery and citeability~\citep{weissburg2024tweets}. Furthermore, because of major commercial stakes with regard to AI and machine learning technology, and e.g.\ following the enthusiastic societal uptake of products employing large language models (LLMs), such as ChatGPT, the pressure to beat competitors as fast as possible is only increasing, and strong eagerness can be observed in many domains to `do something with AI' in order to innovate and remain current.

Where AI used to be a computational modeling tool to better understand human cognition~\citep{vanrooij2023aitheoretical}, the recent interest in AI and LLMs has been turning into one in which AI is seen as a tool that can mimic, surpass and potentially replace human intelligence. In this, the achievement of Artificial General Intelligence (AGI) has become a grand challenge, and in some cases, an explicit business goal. The definition of AGI itself is not as clear-cut or consistent; loosely, it is a phenomenon contrasting with `narrow AI' systems, that were trained for specific tasks~\citep{goertzel2014artificial}. In practice, to demonstrate that the achievement of AGI may be getting closer, researchers have sought to show that AI models generalize to different (and possibly unseen) tasks, with little human intervention, or show performance considered `surprising' to humans.

For example, Google DeepMind claimed their AlphaGeometry model~\citep{trinh2024geometry} reached a `milestone' towards AGI. This model has the ability to solve complex geometry problems, allegedly without the need for human demonstrations during training. However, work such as this had been initially introduced in the 1950s~\citep{hector2024linkedin}: without the use of an LLM, logical inference systems proved 100\% accurate in proving all the theorems of Euclidean Geometry, due to geometry being an axiomatically closed system. Therefore, while DeepMind created a powerfully fast geometry-solving machine, it is still far from AGI.

@@

In light of the encouraging results in Figure~\ref{fig-fomc}, we should expect the probe to predict higher levels of inflation for activations for sentences in the IP category, than for sentences in the DP category. If this was indicative of true intrinsic `understanding' as opposed to memorization, we would not expect to see any significant difference in predicted inflation levels for sentences about birds, independent of whether or not their numbers are increasing. More specifically, we would not expect the probe to predict values for sentences about birds that are substantially different from the values it can be expected to predict for actual white noise.

@@

Linear probes and related tools from mechanistic interpretability were proposed in the context of monitoring models and diagnosing potential problems \citep{alain2018understanding}. Favorable outcomes from probes merely indicate that the model ``has learned information relevant for the property [of interest]" \citep{belinkov2021probing}. Our examples demonstrate that this is achievable even for small models, while these have certainly not developed intrinsic ``understanding'' of the world. Thus, we argue that more conservative and rigorous tests for emerging capabilities of AI model are needed.

Generally, humans are prone to seek patterns everywhere. Meaningful patterns have proven useful in helping us make sense of our past, navigate our present and predict the future. Although this tendency to perceive patterns likely leads to evolutionary benefits even when the perceived patterns are false \citep{foster2009evolution}, psychology has revealed a host of situations in which the ability to perceive patterns severely misfires, leading to irrational beliefs in e.g.\ the power of superstitions \citep{foster2009evolution}, conspiracy theories \citep{van2018connecting}, the paranormal \citep{muller2023linking}, gambler's fallacies \citep{ladouceur1996erroneous} and even interpreting pseudo-profound bullshit as meaningful \citep{walker2019finding}.

We argue herein that AI research and development is a perfect storm that encourages our human biases to perceive spurious sparks of general intelligence in AI systems. When an AI system extracts patterns in the corpus not originally (thought to be) perceived during training, we can easily be misled to perceive and interpret this as the AI system having greater cognitive capabilities. We further elaborate on this by highlighting the risks of finding spurious patterns, and reviewing social science knowledge on the tendency of humans to anthropomorphize and have cognitive bias.

\subsection{Spurious Relationships}
In statistics, misleading patterns are often referred to as spurious relationships: associations, often quantitatively assessed, between two or more variables that are not causally related to each other. Although the formal definition of spuriousness varies somewhat \citep{haig2003spurious}, it distinctly implies that the observation of correlations does not necessarily imply causation. Quantitative data often show non-causal associations (as humorously demonstrated on the \href{http://www.tylervigen.com/spurious-correlations}{Spurious Correlations} website), and as adept as humans are at recognizing patterns, we typically have a much harder time discerning spurious relationships from causal ones.

A major contributor is that humans struggle to tell the difference between random and non-random sequences \citep{falk1997making}, and to generate sequences that appear random \citep{ladouceur1996erroneous}. A common issue is a lack of expectation that randomness that hints towards a causal relationship, such as correlations, will still appear at random. This leads even those trained in statistics and probability to perceive illusory correlations, correlations of inflated magnitude (see \citet{nickerson1998confirmation}), or causal relationships in data that is randomly generated \citep{zgraggen2018investigating}.

\subsection{Anthropomorphism}

Research on anthropomorphism has repeatedly shown the human tendency to attribute human-like characteristics to non-human agents and/or objects. These might include the weather and other natural forces, pets and other animals, gadgets and other pieces of technology~\citep{epley2007seeing}. Formally studied as early as 1944,~\citep{heider1944experimental} observed that humans can correctly interpret a narrative whose characters are abstract 2D shapes, but also that humans interpreted random movements of these shapes as having a human-like narrative. 
Relevant to AI and the degree to which it resembles AGI, anthropomorphizing may occur independently of whether such judgments are accurate, and as a matter of degree: at the weaker end, one may employ anthropomorphism as a metaphorical way of thinking or explaining, and at the stronger end one may attribute human emotions, cognition, and intelligence to AI systems. As \citet{epley2007seeing} note, literature has shown that even weak metaphorical anthropomorphism may affect how humans behave towards non-human agents.

Modern anthropomorphism theory suggests there are three key components, one of which is a cognitive feature, and two of which are motivations. The first involves the easy availability of our experiences as heuristics that can be used to explain external phenomena: ``...knowledge about humans in general, or self-knowledge more specifically, functions as the known and often readily accessible base for induction about the properties of unknown agents" (p.866) \citep{epley2007seeing}. Thus, our experience as humans is an always-readily-available template to interpret the world, including non-human agent behaviors.

This may be more so when the behaviors of that agent are made to resemble humans, which can be a benefit to the second key component of the theory: a motivational state to anthropomorphize among individuals experiencing loneliness, social isolation, or otherwise seeking social connection \citep{epley2007seeing, waytz2010social}.

The third component is the motivation as a human to be competent (effectance motivation). This is most relevant to this discussion, as it describes the need to effectively interact with our environments, including the technologies of the day \citep{epley2007seeing}. When confronted with an opaque technology, a person may interpret its behaviors using the most readily available template at hand, namely their personal human experience, in order to facilitate learning \citep{epley2007seeing, waytz2010social}. Perceiving human characteristics, motivations, emotions, and cognitive processes from one's own experiences in a technology e.g. an AI chatbot, allows for a ready template of comparison at the very least, and possibly an increase in ability to make sense of, and even predict, the agent's behaviors. This may include being placed in a position to master a certain technology, whether by incentives to learn, or fear of poor outcomes should one not manage to learn.

These pressures extend to AI experts, as well as laypersons. In both scholarly and commercial fields, AI experts face considerable pressure to demonstrate competence in their work. Citation metrics and scholarly publications remain the primary metric for tenure and promotion, e.g. \citep{alperin2019significant}, and the number of publications in the AI field has boomed as evidenced by overall (preprint and peer-reviewed) scholarly publications\footnote{\url{https://ourworldindata.org/grapher/annual-scholarly-publications-on-artificial-intelligence?time=2010..2021} }~\citep{Maslej2023-pi}.

The adoption of techniques underlying technologies with the AI label, i.e.\ machine learning, has spread to fields beyond Computer Science, e.g.\ Astronomy, Physics, Medicine and Psychology\footnote{Retrieved 23/01/23 using the search string "TITLE-ABS-KEY ( ( machine  AND  learning )  OR  ( artificial  AND  intelligence )  OR  ai )  AND  PUBYEAR  $>$  2009  AND  PUBYEAR  $<$  2024 " from the \href{https://www.scopus.com/}{SCOPUS} database}. Outside of academia, the number of jobs requiring AI expertise increases rapidly, with demand for `Machine Learning' skills clusters having increased over 500\% from 2010 to 2020 \citep{Maslej2023-pi}. Thus, according to theory, the pressure to demonstrate AI-competence is fertile ground for anthropomorphism to occur.

@@

Confirmation bias is generally defined as favoring interpretations of evidence that support existing beliefs or hypotheses \citep{nickerson1998confirmation}. Theory suggests that it is a category of implicit and unconscious processes that involve assembling one-sided evidence, and shaping it to fit one's belief. Equally important is that theory suggests these behaviors may be motivated or unmotivated, as one may selectively seek evidence in favor of a hypothesis, which one may or may not have a personal interest in supporting.

Hypotheses in present-day AI research are often implicit. Generally, these hypotheses are framed simply as a system being more accurate or efficient, compared to other systems. Where other fields, such as medicine or quantitative social sciences, would further articulate expectations in e.g.\ assigning specific conditions and considering effect sizes assigned to each competing hypothesis, in computer science and AI this is typically not done. This also may have to do with much of the published work being more of an engineering achievement, rather than a true hypothesis test seeking to explain and understand the world. 
However, in discussions on emerging qualities like AGI, this engineering positioning gets muddier, and more formal hypothesis testing would be justifiable: either one interprets outputs as in support of hints towards AGI (the competing hypothesis), or as merely the result of an algorithm integrating qualities from the data it was trained on (the null hypothesis).

Confirmation bias in hypothesis testing may manifest as a number of behaviors (\citet{nickerson1998confirmation}). Scientists may pay little to no attention to competing hypotheses or explanations, e.g. only considering the likelihood that outputs of a system support one's claims, and not the likelihood that the same outputs might occur if one's hypothesis is false. Similarly, bias may show when failing to articulate a sufficiently strong null hypothesis leading to a `weak' or `non-risky' experiment, a problem articulated in response to a number of scientific crises \citep{claesen2022severity}. In extreme cases, propositions may be made that cannot be falsified based on how they are formulated. If the threshold to accept a favored hypothesis is too low, observations consistent with the hypothesis are almost guaranteed, and in turn fail to severely test the claim in question. Thus, one is far more likely to show evidence in favor of their beliefs by posing weak null hypotheses.
